---
title: "Solar Geolocation data on Movebank"
author: "Eldar Rakhimberdiev"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solar Geolocation data on Movebank}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Solar geolocation data on Movebank

Movebank, a depository specialised on the movement data, has a specialised R package [move](https://cran.r-project.org/web/packages/move/index.html).

We will use this package to explore solar geolocation data stored there.

### 0. Setting up

``` {r}
install.packages('move')
library(move)
```

Movebank allows users to see the data only when they are logged in, so one needs to enter his/her credentials otherwise the rest of the script will not work.

```{r eval=F}
loginStored <- movebankLogin(username="user", password="password")
```

### 1. What is available?

Let's find out how many solar geolcoaiton studies are available on Movebank. First we figure out what are the sensor typers we need to look for.

```
getMovebankSensors(login=loginStored)
```
The geolocator data have sensor types 9301403 for raw geolocator data and 3886361 for processed data. How many studies in total we can get to?

```{r}
All1<-getMovebank("study", login=loginStored)
nrow(All1) 
``` 
2577 studies at the moment (2018-09-13)

Now for each study I will get sensors that it used. The following loop goes through all the studies you have permission to see and checks whether the tag type is one of two - 'solar geolocation raw' or 'solar geolocation'.

```{r result=FALSE} 
Res<-c()
Geolocator_studies<-c()
for ( i in 1:nrow(All1)) {
   Try<-getMovebankSensors(study=All1$id[i], login=loginStored)
   if (nrow(Try)>0) {
      if (9301403 %in% unique(Try[,2]) | 3886361 %in% unique(Try[,2])) {
	     Geolocator_studies<-c(Geolocator_studies, All1$id[i])
	     Res<-rbind(Res, Try[which(Try[,2] %in% c(9301403, 3886361)),])
		 	cat('!!\n')
	  }
   }
cat('\r', i)
}

print(length(Geolocator_studies))

print(table(Res$sensor_type_id))
```

At the moment we have only 25 geolocator studies on movebank so far and they have 353 raw data files and 480 processed results with coordinates.

### 2. Download some data

First we check which studies we can actually see.

```{r}
Downloadable_studies<-subset(All1, All1$id %in% Geolocator_studies &  All1$i_can_see_data =='true')
nrow(Downloadable_studies)
```

Now we download data from a sample study with id 242477586. First check the license on this study.
```{r}
print(All1[All1$id==242477586,]$license_terms)
print(All1[All1$id==242477586,]$principal_investigator_name)
```

How many tags are in this study?

```{r}
Barn_swallows<-getMovebankSensors(study=242477586, login=loginStored)
nrow(Barn_swallows)
head(Barn_swallows)
```
There are 18 tags in this study. We now pick the first one and downlaod it.
```{r}
The_swallow <- getMovebankNonLocationData(study=242477586 , sensorID=9301403,
               tag_id=242495280 , login=loginStored)
# or we can do the same by the animal name.
The_swallow <- getMovebankNonLocationData(study=242477586 , sensorID=9301403,
               animalName='H974' , login=loginStored)
			   
head(The_swallow)			   
```
These data are not annotated, but can be saved as .csv and reused.

We can also download all the data from a study with the following line:
```{r}
# Note - ~10 min to run
All_data<-getDataRepositoryData("http://dx.doi.org/10.5441/001/1.rt00m81v")
```
After download data can be plotted or reanalysed.

P.S. The html file from this markdown can be rebuilt with following code
```{r eval=FALSE}
download.file('https://git.io/fAP3J', 'Movebank_geolocaiton_vignette.rmd', cacheOK = FALSE)
rmarkdown::render('Movebank_geolocaiton_vignette.rmd', output_format = 'html_document',
        output_options=list(toc=TRUE, toc_float=list(collapsed=FALSE)), 
        encoding='utf-8')
```